---
title: "Build Population Estimate Data - SQL"
author: "Christopher Prener, Ph.D, Carter Hanford, M.A"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates the requested population estimates for St. Louis County and St. Louis City and passes the data to a SQL database. This notebook pulls tract level data from [Social Explorer](http://socialexplorer.com) and merges it with census tract-level shapefiles from the IPUMS' [NHGIS](https://www.nhgis.org) database to create population estimates by census tract for St. Louis County and St. Louis City (1940-2017).

## Dependencies
This notebook requires a number of different `R` packages:

```{r load-packages}
# tidyverse packages
library(dplyr)         # data wrangling
library(readr)         # working with csv data
library(stringr)       # string tools
library(tidyr)         # data manipulation

# spatial packages
library(areal)         # interpolation
library(sf)            # working with spatial data
library(tidycensus)    # census api access
library(tigris)        # tiger/line api access

# other packages
library(here)          # file path management
library(testthat)      # unit testing
```

We also use a function for unit testing ID numbers:

```{r load-functions}
# load function
source(here("source", "unique_id.R"))
```

We'll also go ahead and load the 2010 census tract shapefile for interpolation:

```{r tracts}
# read in 2010 census tract
stl_tract10 <- tracts(29, county = c(510, 189), year = 2010, class = "sf")
```


### 1940

```{r pop1940}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0012_ds76_1940_tract.csv")) -> pop40

# clean up raw data
pop40 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -PRETRACTA, -AREANAME, -BUQ001, -BUQ002) %>%
  unite(TRACTID, TRACTA, POSTTRCTA, sep ="", na.rm = FALSE) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTID)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop40 = BUB001) %>%
  mutate(TRACTID = str_pad(string = TRACTID, width = 5, side = "left", pad = "0")) -> pop40 
  
# clean data
pop40 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTID, countyID, pop40, GISJOIN) %>%
  mutate(TRACTID = str_replace(string = TRACTID, pattern = "NA$", replacement = "")) %>%
  mutate(countyID = str_replace(string = countyID, pattern = "NA$", replacement = "")) %>%
  rename(tractID = TRACTID) -> pop40
  
# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts40", "STL_BOUNDARY_1940_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl40

# join data to census tract shapefile
stlpop40 <- left_join(stl40, pop40, by = "GISJOIN")
stlpop40 %>%
  select(year, county, tractID, countyID, pop40, geometry) -> stlpop40

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop40, sid = countyID,
                 weight = "total", output = "tibble", 
                 extensive = "pop40") -> tract40

# clean up data frame
tract40$year <- rep(1940, nrow(tract40)) # adds a year column
tract40 %>%
  select(year, GEOID10, pop40) %>%
  rename(GEOID = GEOID10,
         population = pop40) %>%
  na.omit() -> pop40
pop40$population <- round(pop40$population, 0)

# clean up global environment
rm(stl40, stlpop40, tract40)
```

### 1950

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0013_ds82_1950_tract.csv")) -> pop50

# clean up data
pop50 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -PRETRACTA, -AREANAME, -B0J001, -B0J002, -B0J003) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop50 = BZ8001) %>%
  unite(TRACTID, TRACTA, POSTTRCTA, sep ="", na.rm = FALSE) %>%
  mutate(TRACTID = str_pad(string = TRACTID, width = 5, side = "left", pad = "0")) -> pop50 
  
# clean
pop50 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTID, countyID, pop50, GISJOIN) %>%
  mutate(TRACTID = str_replace(string = TRACTID, pattern = "NA$", replacement = "")) %>%
  rename(tractID = TRACTID) -> pop50

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts50", "STL_BOUNDARY_1950_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl50

# join data to census tract shapefile
stl50 <- left_join(stl50, pop50, by = "GISJOIN")
stl50 %>%
  select(year, county, tractID, countyID, pop50, geometry) -> stlpop50

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop50, sid = countyID,
                 weight = "total", output = "tibble", 
                 extensive = "pop50") -> tract50

# clean up data frame
tract50$year <- rep(1950, nrow(tract50)) # adds a year column
tract50 %>%
  select(year, GEOID10, pop50) %>%
  rename(GEOID = GEOID10,
         population = pop50) %>%
  na.omit() -> pop50
pop50$population <- round(pop50$population, 0)

# clean up environment
rm(stl50, stlpop50, tract50)
```

### 1960

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0014_ds92_1960_tract.csv")) -> pop60

# clean up data
pop60 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -PRETRACTA, -AREANAME, -B7B001, -B7B002, -B7B003) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop60 = B53001) %>%
  unite(TRACTID, TRACTA, POSTTRCTA, sep ="", na.rm = FALSE) %>%
  mutate(TRACTID = str_pad(string = TRACTID, width = 5, side = "left", pad = "0")) -> pop60 
  
# clean
pop60 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTID, countyID, pop60, GISJOIN) %>%
  mutate(TRACTID = str_replace(string = TRACTID, pattern = "NA$", replacement = "")) %>%
  rename(tractID = TRACTID) -> pop60

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts60", "STL_BOUNDARY_1960_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl60

# join data to census tract shapefile
stl60 <- left_join(stl60, pop60, by = "GISJOIN")
stl60 %>%
  select(year, county, tractID, countyID, pop60, geometry) -> stlpop60

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop60, sid = countyID,
                 weight = "sum", output = "tibble", 
                 extensive = "pop60") -> tract60

# clean up data frame
tract60$year <- rep(1960, nrow(tract60)) # adds a year column
tract60 %>%
  select(year, GEOID10, pop60) %>%
  rename(GEOID = GEOID10,
         population = pop60) %>%
  na.omit() -> pop60
pop60$population <- round(pop60$population, 0)

# clean up environment
rm(stl60, stlpop60, tract60)
```

### 1970

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0015_ds98_1970_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") -> pop70

# clean up data
pop70 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -AREANAME, -CTY_SUBA,
         -PLACEA, -SCSAA, -URB_AREAA, -BLOCKA, -CDA, -SMSAA, -C0X001, -C0X002, -C0X003) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop70 = C1I001) -> pop70 
  
# clean
pop70 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop70, GISJOIN) %>%
  rename(tractID = TRACTA) -> pop70

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts70", "STL_BOUNDARY_1970_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl70

# join data to census tract shapefile
stl70 <- left_join(stl70, pop70, by = "GISJOIN")
stl70 %>%
  select(year, county, tractID, countyID, pop70, geometry) -> stlpop70

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop70, sid = countyID,
                 weight = "sum", output = "tibble", 
                 extensive = "pop70") -> tract70

# clean up data frame
tract70$year <- rep(1970, nrow(tract70)) # adds a year column
tract70 %>%
  select(year, GEOID10, pop70) %>%
   rename(GEOID = GEOID10,
         population = pop70) %>%
  na.omit() -> pop70
pop70$population <- round(pop70$population, 0)

# clean up environment
rm(stl70, stlpop70, tract70)
```


### 1980

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0016_ds116_1980_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") -> pop80

# clean up data
pop80 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -AREANAME, -CTY_SUBA,
         -PLACEA, -BLOCKA, -C6X001, -C6X002, -C6X003, -C6X004, -C6X005) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop80 = C6W001) -> pop80 
  
# clean
pop80 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop80, GISJOIN) %>%
  rename(tractID = TRACTA) -> pop80

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts80", "STL_BOUNDARY_1980_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl80

# join data to census tract shapefile
stl80 <- left_join(stl80, pop80, by = "GISJOIN")
stl80 %>%
  select(year, county, tractID, countyID, pop80, geometry) -> stlpop80

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop80, sid = countyID,
                 weight = "sum", output = "tibble", 
                 extensive = "pop80") -> tract80

# clean up data frame
tract80$year <- rep(1980, nrow(tract80)) # adds a year column
tract80 %>%
  select(year, GEOID10, pop80) %>%
   rename(GEOID = GEOID10,
         population = pop80) %>%
  na.omit() -> pop80
pop80$population <- round(pop80$population, 0)

# clean up environment
rm(stl80, stlpop80, tract80)
```

### 1990

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0017_ds120_1990_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") -> pop90

# clean up data
pop90 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -CTY_SUBA,
         -PLACEA, -BLOCKA, -ANRCA, -AIANHHA, -RES_ONLYA, -RES_TRSTA,
         -TRUSTA, -BLCK_GRPA, -CDA, -C_CITYA, -DIVISIONA, -MSA_CMSAA, 
         -PMSAA, -REGIONA, -URBRURALA, -URB_AREAA, 
         -CD103A, -ANPSADPI, -EUY001, -EUY002, -EUY003, -EUY004, -EUY005) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop90 = ET1001) -> pop90 
  
# clean
pop90 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop90, GISJOIN) %>%
  rename(tractID = TRACTA) -> pop90

# read in 1990 census counts, clean
get_decennial(geography = "tract", variable = "P0010001", year = 1990, state = 29, county = 510, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value) %>%
  rename(pop90 = value) -> stlcity90

get_decennial(geography = "tract", variable = "P0010001", year = 1990, state = 29, county = 189, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value) %>%
  rename(pop90 = value) -> stlcounty90

rbind(stlcity90, stlcounty90) -> stlpop90
stlpop90 <- left_join(stlpop90, pop90, by = "pop90")

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop90, sid = countyID,
                 weight = "sum", output = "tibble", 
                 extensive = "pop90") -> tract90

# clean up data frame
tract90$year <- rep(1990, nrow(tract90)) # adds a year column
tract90 %>%
  select(year, GEOID10, pop90) %>%
  rename(GEOID = GEOID10,
         population = pop90) %>%
  na.omit() -> pop90
pop90$population <- round(pop90$population, 0)

# clean up environment
rm(stl80, stlpop80, tract80, stlcity90, stlcounty90)
```

### 2000

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0021_ds146_2000_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") -> pop00

# clean up data
pop00 %>%
  select(-STATE, -STATEA, -GISJOIN, -CTY_SUBA,
         -PLACEA, -ANRCA, -AIANHHA, -RES_ONLYA,
         -TRUSTA, -BLCK_GRPA, -C_CITYA, -DIVISIONA, -MSA_CMSAA, 
         -PMSAA, -REGIONA, -URB_AREAA, -TRBL_CTA, -TRBL_BGA, -TRBL_SUBA,
         -NECMAA, -CD106A, -CD108A, -CD109A, -ZIP3A, -ZCTAA, -NAME) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop00 = FL5001) -> pop00 

# clean
pop00 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop00) %>%
  rename(tractID = TRACTA) -> pop00

# read in 2000 census counts, clean
get_decennial(geography = "tract", variable = "P001001", year = 2000, state = 29, county = 510, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value) %>%
  rename(pop00 = value) -> stlcity00

get_decennial(geography = "tract", variable = "P001001", year = 2000, state = 29, county = 189, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value) %>%
  rename(pop00 = value) -> stlcounty00

rbind(stlcity00, stlcounty00) -> stlpop00
stlpop00 <- left_join(stlpop00, pop00, by = "pop00")

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlpop00, sid = countyID,
                 weight = "sum", output = "tibble", 
                 extensive = "pop00") -> tract00

# clean up data frame
tract00$year <- rep(2000, nrow(tract00)) # adds a year column
tract00 %>%
  select(year, GEOID10, pop00) %>%
  rename(GEOID = GEOID10,
         population = pop00) %>%
  na.omit() -> pop00
pop00$population <- round(pop00$population, 0)

# clean up environment
rm(stl00, stlpop00, tract00, stlcity00, stlcounty00)
```

### 2010

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0022_ds172_2010_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop10

# clean up data
pop10 %>%
  select(-GISJOIN, -STATE, -STATEA, -PLACEA, -ANRCA, -AIANHHA, -RES_ONLYA,
         -TRUSTA, -DIVISIONA, -REGIONA, -NAME, -COUSUBA, -BLKGRPA, -BLOCKA, -CONCITA, -AITSCEA,
          -TTRACTA, -TBLKGRPA, -CBSAA, -METDIVA, -CSAA, -NECTAA, -NECTADIVA, -CNECTAA, -UAA, -URBRURALA, 
         -CDA, -SLDUA, -SLDLA, -ZCTA5A, -SUBMCDA, -SDELMA, -SDSECA, -SDUNIA, -SABINSA) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop10 = H7V001) -> pop10

# clean
pop10 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop10) %>%
  rename(tractID = TRACTA) -> pop10

get_decennial(geography = "tract", variable = "P001001", year = 2010, state = 29, county = 510, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value) %>%
  rename(pop10 = value) -> stlcity10

get_decennial(geography = "tract", variable = "P001001", year = 2010, state = 29, county = 189, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value) %>%
  rename(pop10 = value) -> stlcounty10

rbind(stlcity10, stlcounty10) -> stlpop10
stlpop10 <- left_join(stlpop10, pop10, by = "pop10")

# clean up data frame
st_geometry(stlpop10) <- NULL
stlpop10 %>%
  select(year, GEOID, pop10) %>%
  rename(population = pop10) -> pop10

# clean up environment
rm(stlpop10, stlcity10, stlcounty10)
```

### 2011

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0023_ds184_20115_2011_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop11

# clean up data
pop11 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, MNTE001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop11 = MNTE001) -> pop11 

# clean
pop11 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop11) %>%
  rename(tractID = TRACTA) -> pop11

# remove excess years in year column
str_replace_all(pop11$year, "2007-", "") -> pop11$year

# read in 2011 census counts
get_acs(geography = "tract", year = 2011, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop11 = estimate,
         pop11_m = moe) -> stlcity11

get_acs(geography = "tract", year = 2011, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop11 = estimate,
         pop11_m = moe) -> stlcounty11

rbind(stlcity11, stlcounty11) -> stlpop11
stlpop11 <- left_join(stlpop11, pop11, by = "pop11")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop11 %>%
  select(year, GEOID, pop11) %>%
  rename(population = pop11) -> pop11

# clean up environment
rm(stlpop11, stlcity11, stlcounty11)
```

### 2012

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0024_ds191_20125_2012_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop12

# clean up data
pop12 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, QSPE001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop12 = QSPE001) -> pop12 

# clean
pop12 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop12) %>%
  rename(tractID = TRACTA) -> pop12

# remove excess years in year column
str_replace_all(pop12$year, "2008-", "") -> pop12$year

# read in 2012 census counts
get_acs(geography = "tract", year = 2012, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop12 = estimate,
         pop12_m = moe) -> stlcity12

get_acs(geography = "tract", year = 2012, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop12 = estimate,
         pop12_m = moe) -> stlcounty12

rbind(stlcity12, stlcounty12) -> stlpop12
stlpop12 <- left_join(stlpop12, pop12, by = "pop12")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop12 %>%
  select(year, GEOID, pop12) %>% 
  rename(population = pop12) -> pop12

# clean up environment
rm(stlpop12, stlcity12, stlcounty12)
```

### 2013

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0025_ds201_20135_2013_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop13

# clean up data
pop13 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, UEPE001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop13 = UEPE001) -> pop13 

# clean
pop13 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop13) %>%
  rename(tractID = TRACTA) -> pop13

# remove excess years in year column
str_replace_all(pop13$year, "2009-", "") -> pop13$year

# read in 2013 census counts
get_acs(geography = "tract", year = 2013, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop13 = estimate,
         pop13_m = moe) -> stlcity13

get_acs(geography = "tract", year = 2013, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop13 = estimate,
         pop13_m = moe) -> stlcounty13

rbind(stlcity13, stlcounty13) -> stlpop13
stlpop13 <- left_join(stlpop13, pop13, by = "pop13")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop13 %>%
  select(year, GEOID, pop13) %>%
  rename(population = pop13) -> pop13

# clean up environment
rm(stlpop13, stlcity13, stlcounty13)
```

### 2014

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0026_ds206_20145_2014_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop14

# clean up data
pop14 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, ABA1E001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop14 = ABA1E001) -> pop14

# clean
pop14 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop14) %>%
  rename(tractID = TRACTA) -> pop14

# remove excess years in year column
str_replace_all(pop14$year, "2010-", "") -> pop14$year

# read in 2014 census counts
get_acs(geography = "tract", year = 2014, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop14 = estimate,
         pop14_m = moe) -> stlcity14

get_acs(geography = "tract", year = 2014, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop14 = estimate,
         pop14_m = moe) -> stlcounty14

rbind(stlcity14, stlcounty14) -> stlpop14
stlpop14 <- left_join(stlpop14, pop14, by = "pop14")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop14 %>%
  select(year, GEOID, pop14) %>%
  rename(population = pop14) -> pop14

# clean up environment
rm(stlpop14, stlcity14, stlcounty14)
```

### 2015

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0027_ds215_20155_2015_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop15

# clean up data
pop15 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, ADKWE001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop15 = ADKWE001) -> pop15

# clean
pop15 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop15) %>%
  rename(tractID = TRACTA) -> pop15

# remove excess years in year column
str_replace_all(pop15$year, "2011-", "") -> pop15$year

# read in 2015 census counts
get_acs(geography = "tract", year = 2015, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop15 = estimate,
         pop15_m = moe) -> stlcity15

get_acs(geography = "tract", year = 2015, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop15 = estimate,
         pop15_m = moe) -> stlcounty15

rbind(stlcity15, stlcounty15) -> stlpop15
stlpop15 <- left_join(stlpop15, pop15, by = "pop15")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop15 %>%
  select(year, GEOID, pop15) %>%
  rename(population = pop15) -> pop15

# clean up environment
rm(stlpop15, stlcity15, stlcounty15)
```

### 2016

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0028_ds225_20165_2016_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop16

# clean up data
pop16 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, AF2LE001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop16 = AF2LE001) -> pop16

# clean
pop16 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop16) %>%
  rename(tractID = TRACTA) -> pop16

# remove excess years in year column
str_replace_all(pop16$year, "2012-", "") -> pop16$year

# read in 2016 census counts
get_acs(geography = "tract", year = 2016, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop16 = estimate,
         pop16_m = moe) -> stlcity16

get_acs(geography = "tract", year = 2016, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop16 = estimate,
         pop16_m = moe) -> stlcounty16

rbind(stlcity16, stlcounty16) -> stlpop16
stlpop16 <- left_join(stlpop16, pop16, by = "pop16")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop16 %>%
  select(year, GEOID, pop16) %>%
  rename(population = pop16) -> pop16

# clean up environment
rm(stlpop16, stlcity16, stlcounty16)
```

### 2017

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "nhgis0029_ds233_20175_2017_tract.csv")) %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St. Louis County" | COUNTY == "St. Louis city") -> pop17

# clean up data
pop17 %>%
  select(YEAR, COUNTYA, COUNTY, TRACTA, AHY1E001) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         pop17 = AHY1E001) -> pop17

# clean
pop17 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, pop17) %>%
  rename(tractID = TRACTA) -> pop17

# remove excess years in year column
str_replace_all(pop17$year, "2013-", "") -> pop17$year

# read in 2017 census counts
get_acs(geography = "tract", year = 2017, variables = "B01003_001", state = 29, county = 510) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop17 = estimate,
         pop17_m = moe) -> stlcity17

get_acs(geography = "tract", year = 2017, variables = "B01003_001", state = 29, county = 189) %>%
  select(GEOID, estimate, moe) %>%
  rename(pop17 = estimate,
         pop17_m = moe) -> stlcounty17

rbind(stlcity17, stlcounty17) -> stlpop17
stlpop17 <- left_join(stlpop17, pop17, by = "pop17")

# clean up data frame
#st_geometry(stlpop11) <- NULL
stlpop17 %>%
  select(year, GEOID, pop17) %>%
  rename(population = pop17) -> pop17

# clean up environment
rm(stlpop17, stlcity17, stlcounty17)
```

### Combine data

```{r}
# combine all data to a single dataframe
rbind(pop40, pop50, 
      pop60, pop80, 
      pop90, pop00, 
      pop10, pop11, 
      pop12, pop13, 
      pop14, pop15, 
      pop16, pop17) -> stl_population
```
