---
title: "Build Race Estimate Data - SQL"
author: "Christopher Prener, Ph.D, Carter Hanford, M.A"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates race estimates for St. Louis County and St. Louis City and passes the data to a SQL database. This notebook pulls tract level data from [Social Explorer](http://socialexplorer.com) and merges it with census tract-level shapefiles from the IPUMS' [NHGIS](https://www.nhgis.org) database to create race estimates by census tract for St. Louis County and St. Louis City (1940-2017).

## Dependencies
This notebook requires a number of different `R` packages:

```{r load-packages, results = 'hide'}
# tidyverse packages
library(dplyr)         # data wrangling
library(readr)         # working with csv data
library(stringr)       # string tools
library(tidyr)         # data manipulation

# spatial packages
library(areal)         # interpolation
library(sf)            # working with spatial data
library(tidycensus)    # census api access
library(tigris)        # tiger/line api access

# other packages
library(here)          # file path management
library(testthat)      # unit testing
```

We also use a function for unit testing ID numbers:

```{r load-functions}
# load function
source(here("source", "unique_id.R"))
```

We'll also go ahead and load the 2010 census tract shapefile for interpolation:

```{r tracts}
# read in 2010 census tract
stl_tract10 <- tracts(29, county = c(510, 189), year = 2010, class = "sf")
```

### 1940 

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "race", "nhgis0030_ds76_1940_tract.csv")) -> race40

# clean up raw data
race40 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -PRETRACTA, -AREANAME) %>%
  unite(TRACTID, TRACTA, POSTTRCTA, sep ="", na.rm = FALSE) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTID)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         white40 = BUQ001,
         black40 = BUQ002) %>%
  mutate(TRACTID = str_pad(string = TRACTID, width = 5, side = "left", pad = "0")) -> race40 

# clean data
race40 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTID, countyID, white40, black40, GISJOIN) %>%
  mutate(TRACTID = str_replace(string = TRACTID, pattern = "NA$", replacement = "")) %>%
  mutate(countyID = str_replace(string = countyID, pattern = "NA$", replacement = "")) %>%
  rename(tractID = TRACTID) -> race40

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts40", "STL_BOUNDARY_1940_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl40

# join data to census tract shapefile
stlrace40 <- left_join(stl40, race40, by = "GISJOIN")
stlrace40 %>%
  select(year, county, tractID, countyID, white40, black40, geometry) -> stlrace40

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlrace40, sid = tractID,
                 weight = "sum", output = "tibble", 
                 extensive = "white40") -> tract_white40

stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlrace40, sid = countyID,
                 weight = "total", output = "tibble", 
                 extensive = "black40") -> tract_black40

# clean up data frame
tract_white40 %>%
  select(GEOID10, white40) %>%
  rename(GEOID = GEOID10,
         white = white40) %>%
  na.omit() -> white40

tract_black40 %>%
  select(GEOID10, black40) %>%
  rename(GEOID = GEOID10,
         black = black40) %>%
  na.omit() -> black40

race40 <- left_join(white40, black40, by = "GEOID")
race40 %>%
  select(GEOID, white, black) -> race40

# pivot race column from wide to long
race40 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race40
race40$year <- rep(1940, nrow(race40)) # adds a year column
race40 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race40

# clean up environment
rm(tract_white40, tract_black40, stl40, stlrace40, white40, black40)
```

### 1950 

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "race", "nhgis0030_ds82_1950_tract.csv")) -> race50

# clean up raw data
race50 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -PRETRACTA, -AREANAME) %>%
  unite(TRACTID, TRACTA, POSTTRCTA, sep ="", na.rm = FALSE) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTID)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         white50 = B0J001,
         black50 = B0J002) %>%
  mutate(TRACTID = str_pad(string = TRACTID, width = 5, side = "left", pad = "0")) -> race50 

# clean data
race50 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTID, countyID, white50, black50, GISJOIN) %>%
  mutate(TRACTID = str_replace(string = TRACTID, pattern = "NA$", replacement = "")) %>%
  mutate(countyID = str_replace(string = countyID, pattern = "NA$", replacement = "")) %>%
  rename(tractID = TRACTID) -> race50

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts50", "STL_BOUNDARY_1950_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl50

# join data to census tract shapefile
stlrace50 <- left_join(stl50, race50, by = "GISJOIN")
stlrace50 %>%
  select(year, county, tractID, countyID, white50, black50, geometry) -> stlrace50

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlrace50, sid = tractID,
                 weight = "sum", output = "tibble", 
                 extensive = c("white50", "black50")) -> tract50

# clean up data frame
tract50 %>%
  select(GEOID10, white50, black50) %>%
  rename(GEOID = GEOID10,
         white = white50,
         black = black50) %>%
  na.omit() -> race50

# pivot race column from wide to long
race50 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race50
race50$year <- rep(1950, nrow(race50)) # adds a year column
race50 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race50

# clean up environment
rm(tract50, stl50, stlrace50)
```

### 1960 

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "race", "nhgis0030_ds92_1960_tract.csv")) -> race60

# clean up raw data
race60 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -PRETRACTA, -AREANAME) %>%
  unite(TRACTID, TRACTA, POSTTRCTA, sep ="", na.rm = FALSE) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTID)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         white60 = B7B001,
         black60 = B7B002) %>%
  mutate(TRACTID = str_pad(string = TRACTID, width = 5, side = "left", pad = "0")) -> race60 

# clean data
race60 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTID, countyID, white60, black60, GISJOIN) %>%
  mutate(TRACTID = str_replace(string = TRACTID, pattern = "NA$", replacement = "")) %>%
  mutate(countyID = str_replace(string = countyID, pattern = "NA$", replacement = "")) %>%
  rename(tractID = TRACTID) -> race60

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts60", "STL_BOUNDARY_1960_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl60

# join data to census tract shapefile
stlrace60 <- left_join(stl60, race60, by = "GISJOIN")
stlrace60 %>%
  select(year, county, tractID, countyID, white60, black60, geometry) -> stlrace60

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlrace60, sid = tractID,
                 weight = "sum", output = "tibble", 
                 extensive = c("white60", "black60")) -> tract60

# clean up data frame
tract60 %>%
  select(GEOID10, white60, black60) %>%
  rename(GEOID = GEOID10,
         white = white60,
         black = black60) %>%
  na.omit() -> race60

# pivot race column from wide to long
race60 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race60
race60$year <- rep(1960, nrow(race60)) # adds a year column
race60 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race60

# clean up environment
rm(tract60, stl60, stlrace60)
```

### 1970 

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "race", "nhgis0030_ds98_1970_tract.csv")) -> race70

# clean up raw data
race70 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -AREANAME, -CTY_SUBA,
         -PLACEA, -SCSAA, -SMSAA, -URB_AREAA, -BLOCKA, -CDA) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>%
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         white70 = C0X001,
         black70 = C0X002) -> race70 

# clean
race70 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, white70, black70, GISJOIN) %>%
  rename(tractID = TRACTA) -> race70

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts70", "STL_BOUNDARY_1970_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl70

# join data to census tract shapefile
stlrace70 <- left_join(stl70, race70, by = "GISJOIN")
stlrace70 %>%
  select(year, county, tractID, countyID, white70, black70, geometry) -> stlrace70

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlrace70, sid = tractID,
                 weight = "sum", output = "tibble", 
                 extensive = c("white70", 
                               "black70")) -> tract70

# clean up data frame
tract70 %>%
  select(GEOID10, white70, black70) %>%
  rename(GEOID = GEOID10,
         white = white70,
         black = black70) %>%
  na.omit() -> race70

# pivot race column from wide to long
race70 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race70
race70$year <- rep(1970, nrow(race70)) # adds a year column
race70 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race70

# clean up environment
rm(tract70, stl70, stlrace70)
```

### 1980 

```{r}
# read in tract level population estimates
read_csv(here("data", "raw", "IPUMS", "race", "nhgis0030_ds104_1980_tract.csv")) -> race80

# clean up raw data
race80 %>%
  filter(STATE == "Missouri") %>%
  filter(COUNTY == "St Louis" | COUNTY == "St Louis City") %>%
  select(-STATE, -STATEA, -AREANAME, -CTY_SUBA,
         -PLACEA, -SCSAA, -SMSAA, -URB_AREAA, -BLOCKA, -CDA) %>%
  mutate(county = paste0(COUNTYA, "-", TRACTA)) %>% 
  rename(countyID = county,
         year = YEAR,
         county = COUNTY,
         white80 = C9D001,
         black80 = C9D002) -> race80 

# clean
race80 %>%
  select(-COUNTYA) %>%
  select(year, county, TRACTA, countyID, white80, black80, GISJOIN) %>%
  rename(tractID = TRACTA) -> race80

# read in census tract shapefile
st_read(here("data", "spatial", "STL_DEMOGRAPHICS_tracts80", "STL_BOUNDARY_1980_tracts.geojson"),
        stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) -> stl80

# join data to census tract shapefile
stlrace80 <- left_join(stl80, race80, by = "GISJOIN")
stlrace80 %>%
  select(year, county, tractID, countyID, white80, black80, geometry) -> stlrace80

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stlrace80, sid = tractID,
                 weight = "sum", output = "tibble", 
                 extensive = c("white80", 
                               "black80")) -> tract80

# clean up data frame
tract80 %>%
  select(GEOID10, white80, black80) %>%
  rename(GEOID = GEOID10,
         white = white80,
         black = black80) %>%
  na.omit() -> race80

# pivot race column from wide to long
race80 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race80
race80$year <- rep(1980, nrow(race80)) # adds a year column
race80 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race80

# clean up environment
rm(tract80, stl80, stlrace80)
```

### 1990 

```{r}
# read in tract level race estimates - white
get_decennial(geography = "tract", variables = "P0060001", state = 29, county = c(510, 189), year = 1990, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value, NAME) %>%
  rename(white90 = value,
         tract = NAME) -> white90

# read in tract level race estimates - black
get_decennial(geography = "tract", variables = "P0060002", state = 29, county = c(510, 189), year = 1990, geometry = FALSE) %>% 
  select(GEOID, value, NAME) %>%
  rename(black90 = value,
         tract = NAME) -> black90

# combine
stl90 <- left_join(white90, black90, by = "GEOID")
stl90 %>%
  select(-tract.x) %>%
  rename(tract = tract.y) -> stl90

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stl90, sid = tract,
                 weight = "sum", output = "tibble", 
                 extensive = c("white90", 
                               "black90")) -> tract90

# clean up data frame
tract90 %>%
  select(GEOID10, white90, black90) %>%
  rename(GEOID = GEOID10,
         white = white90,
         black = black90) %>%
  na.omit() -> race90

# pivot race column from wide to long
race90 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race90
race90$year <- rep(1990, nrow(race90)) # adds a year column
race90 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race90

# clean up environment
rm(tract90, stl90, stlrace90, stlcity90, stlcounty90, white90, black90)
```

### 2000

```{r}
# read in tract level race estimates - white
get_decennial(geography = "tract", variables = "P003003", state = 29, county = c(510, 189), year = 2000, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value, NAME) %>%
  rename(white00 = value,
         tract = NAME) -> white00

# read in tract level race estimates - black
get_decennial(geography = "tract", variables = "P003004", state = 29, county = c(510, 189), year = 2000, geometry = FALSE) %>% 
  select(GEOID, value, NAME) %>%
  rename(black00 = value,
         tract = NAME) -> black00

# combine
stl00 <- left_join(white00, black00, by = "GEOID")
stl00 %>%
  select(-tract.x) %>%
  rename(tract = tract.y) -> stl00

# interpolate to 2010 census tract
stl_tract10 %>%
  st_transform(crs = 26915) %>%
  select(GEOID10) %>%
  aw_interpolate(tid = GEOID10, source = stl00, sid = tract,
                 weight = "sum", output = "tibble", 
                 extensive = c("white00", 
                               "black00")) -> tract00

# clean up data frame
tract00 %>%
  select(GEOID10, white00, black00) %>%
  rename(GEOID = GEOID10,
         white = white00,
         black = black00) %>%
  na.omit() -> race00

# pivot race column from wide to long
race00 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race00
race00$year <- rep(2000, nrow(race00)) # adds a year column
race00 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race00

# clean up environment
rm(tract00, stl00, stlrace00, white00, black00)
```

### 2010

```{r}
# read in tract level race estimates - white
get_decennial(geography = "tract", variables = "P003002", state = 29, county = c(510, 189), year = 2010, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, value, NAME) %>%
  rename(white10 = value,
         tract = NAME) -> white10

# read in tract level race estimates - black
get_decennial(geography = "tract", variables = "P003003", state = 29, county = c(510, 189), year = 2010, geometry = FALSE) %>% 
  select(GEOID, value, NAME) %>%
  rename(black10 = value,
         tract = NAME) -> black10

# combine
stl10 <- left_join(white10, black10, by = "GEOID")
stl10 %>%
  select(-tract.x) %>%
  rename(tract = tract.y) -> stl10

# clean up data frame
stl10 %>%
  select(GEOID, white10, black10) %>%
  rename(white = white10,
         black = black10) %>%
  na.omit() -> race10
st_geometry(race10) <- NULL

# pivot race column from wide to long
race10 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race10
race10$year <- rep(2010, nrow(race10)) # adds a year column
race10 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race10

# clean up environment
rm(stl10, white10, black10)
```

### 2011

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2011, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white11 = estimate) -> white11

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2011, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black11 = estimate) -> black11

# combine
stl11 <- left_join(white11, black11, by = "GEOID")

# clean up data frame
stl11 %>%
  select(GEOID, white11, black11) %>%
  rename(white = white11,
         black = black11) %>%
  na.omit() -> race11
st_geometry(race11) <- NULL

# pivot race column from wide to long
race11 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race11
race11$year <- rep(2011, nrow(race11)) # adds a year column
race11 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race11

# clean up environment
rm(stl11, white11, black11)
```

### 2012

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2012, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white12 = estimate) -> white12

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2012, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black12 = estimate) -> black12

# combine
stl12 <- left_join(white12, black12, by = "GEOID")

# clean up data frame
stl12 %>%
  select(GEOID, white12, black12) %>%
  rename(white = white12,
         black = black12) %>%
  na.omit() -> race12
st_geometry(race12) <- NULL

# pivot race column from wide to long
race12 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race12
race12$year <- rep(2012, nrow(race12)) # adds a year column
race12 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race12

# clean up environment
rm(stl12, white12, black12)
```

### 2013

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2013, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white13 = estimate) -> white13

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2013, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black13 = estimate) -> black13

# combine
stl13 <- left_join(white13, black13, by = "GEOID")

# clean up data frame
stl13 %>%
  select(GEOID, white13, black13) %>%
  rename(white = white13,
         black = black13) %>%
  na.omit() -> race13
st_geometry(race13) <- NULL

# pivot race column from wide to long
race13 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race13
race13$year <- rep(2013, nrow(race13)) # adds a year column
race13 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race13

# clean up environment
rm(stl13, white13, black13)
```

### 2014

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2014, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white14 = estimate) -> white14

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2014, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black14 = estimate) -> black14

# combine
stl14 <- left_join(white14, black14, by = "GEOID")

# clean up data frame
stl14 %>%
  select(GEOID, white14, black14) %>%
  rename(white = white14,
         black = black14) %>%
  na.omit() -> race14
st_geometry(race14) <- NULL

# pivot race column from wide to long
race14 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race14
race14$year <- rep(2014, nrow(race14)) # adds a year column
race14 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race14

# clean up environment
rm(stl14, white14, black14)
```

### 2015

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2015, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white15 = estimate) -> white15

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2015, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black15 = estimate) -> black15

# combine
stl15 <- left_join(white15, black15, by = "GEOID")

# clean up data frame
stl15 %>%
  select(GEOID, white15, black15) %>%
  rename(white = white15,
         black = black15) %>%
  na.omit() -> race15
st_geometry(race15) <- NULL

# pivot race column from wide to long
race15 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race15
race15$year <- rep(2015, nrow(race15)) # adds a year column
race15 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race15

# clean up environment
rm(stl15, white15, black15)
```

### 2016

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2016, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white16 = estimate) -> white16

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2016, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black16 = estimate) -> black16

# combine
stl16 <- left_join(white16, black16, by = "GEOID")

# clean up data frame
stl16 %>%
  select(GEOID, white16, black16) %>%
  rename(white = white16,
         black = black16) %>%
  na.omit() -> race16
st_geometry(race16) <- NULL

# pivot race column from wide to long
race16 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race16
race16$year <- rep(2016, nrow(race16)) # adds a year column
race16 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race16

# clean up environment
rm(stl16, white16, black16)
```

### 2017

```{r}
# read in tract level race estimates - white
get_acs(geography = "tract", variables = "B02001_002", state = 29, county = c(510, 189), year = 2017, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(white17 = estimate) -> white17

# read in tract level race estimates - black
get_acs(geography = "tract", variables = "B02001_003", state = 29, county = c(510, 189), year = 2017, geometry = FALSE) %>% 
  select(GEOID, estimate) %>%
  rename(black17 = estimate) -> black17

# combine
stl17 <- left_join(white17, black17, by = "GEOID")

# clean up data frame
stl17 %>%
  select(GEOID, white17, black17) %>%
  rename(white = white17,
         black = black17) %>%
  na.omit() -> race17
st_geometry(race17) <- NULL

# pivot race column from wide to long
race17 %>%
  pivot_longer(-GEOID, names_to = "race", values_to = "count") -> race17
race17$year <- rep(2017, nrow(race17)) # adds a year column
race17 %>%
  select(year, GEOID, race, count) %>%
  rename(value = race) -> race17

# clean up environment
rm(stl17, white17, black17)
```

### Combine data

```{r}
# combine all data to a single dataframe
rbind(race40, race50, 
      race60, race80, 
      race90, race00, 
      race10, race11, 
      race12, race13, 
      race14, race15, 
      race16, race17) -> stl_race

stl_race %>%
  rename(estimate = count) ->stl_race
```
